---
title: "Time Series Analysis"
format: html
editor: visual
---

## Introduction

This document contains time series analysis for the DSAN5100 final project.

# Load data files and preprocess

```{r}
# Load required libraries for R
library(tidyverse)
library(lubridate)
library(arrow)
library(plotly)
library(zoo)
library(forecast)  
library(imputeTS)
library(tseries)
library(patchwork)
library(astsa)

# Set Arrow option to skip null characters
options(arrow.skip_nul = TRUE)

# Set seed for reproducibility
set.seed(42)
```

```{r}
# Load the data

argentina_df <- read_parquet("data/argentina_data_final.parquet")
canada_df <- read_parquet("data/canada_data_final.parquet")
china_df <- read_parquet("data/china_data_final.parquet")
india_df <- read_parquet("data/india_data_final.parquet")
italy_df <- read_parquet("data/italy_data_final.parquet")
russia_df <- read_parquet("data/russia_data_final.parquet")
us_df <- read_parquet("data/us_data_final.parquet")


head(us_df)
```

```{r}
# Change publishedAt column into a date type

# Function to convert publishedAt to date type
convert_date_column <- function(df) {
  df <- df %>%
    mutate(publishedAt = as.Date(publishedAt))
  return(df)
}

# Apply date conversion to all dataframes
argentina_df <- convert_date_column(argentina_df)
canada_df <- convert_date_column(canada_df)
china_df <- convert_date_column(china_df)
india_df <- convert_date_column(india_df)
italy_df <- convert_date_column(italy_df)
russia_df <- convert_date_column(russia_df)
us_df <- convert_date_column(us_df)

# Check the data type conversion
cat("Date type conversion completed:\n")
cat("publishedAt column type for US data:", class(us_df$publishedAt), "\n")
```

```{r}
# Check date range for various country data

cat("Date range for US data:", as.character(range(us_df$publishedAt, na.rm = TRUE)), "\n")
cat("Date range for Argentina data:", as.character(range(argentina_df$publishedAt, na.rm = TRUE)), "\n")
cat("Date range for Canada data:", as.character(range(canada_df$publishedAt, na.rm = TRUE)), "\n")
cat("Date range for China data:", as.character(range(china_df$publishedAt, na.rm = TRUE)), "\n")
cat("Date range for India data:", as.character(range(india_df$publishedAt, na.rm = TRUE)), "\n")
cat("Date range for Italy data:", as.character(range(italy_df$publishedAt, na.rm = TRUE)), "\n")
cat("Date range for Russia data:", as.character(range(russia_df$publishedAt, na.rm = TRUE)), "\n")
```

*All of the countries have an end date of 2021-11-29 after which the data was no longer collected. The earliest data for the published articles range from 2010-11-24 for Russia to 2019-11-23 for China*

```{r}
# Combine the country datasets into df_countries

# Set Arrow option to handle null characters
options(arrow.skip_nul = TRUE)

df_countries <- bind_rows(
  us_df, 
  canada_df, 
  china_df, 
  india_df, 
  italy_df, 
  russia_df, 
  argentina_df
)

# Check the combined dataset
cat("Combined dataset dimensions:", nrow(df_countries), "rows,", ncol(df_countries), "columns\n")
cat("Countries included:", unique(df_countries$location), "\n")
```

```{r}
# Filter the dataset to start from 2020 for more focused analysis
df_countries_2020 <- df_countries %>% filter(year > 2019)
```

# Analysis Health, General, Science categories combined

*In to capture the sentiment around COVID-19 and also leverage the scarse related data in the dataset, we have combined the Health, Science, and General category to explore the sentiment and bias during the COVID-19 pandemic (2020-02 to 2021-11)*

```{r}
# Combined time series plot for sentiment and bias across health, general, and science categories, daily

# Prepare daily combined data for USA with both sentiment and bias for health, general, and science categories
usa_combined_daily <- df_countries %>%
  filter(location == "United States", category %in% c("health", "general", "science", "business")) %>%
  group_by(publishedAt) %>%
  summarise(
    avg_sentiment = mean(compound, na.rm = TRUE),
    avg_bias = mean(bias_score, na.rm = TRUE),
    n_articles = n(),
    .groups = 'drop'
  ) %>%
  rename(date = publishedAt) %>%
  # Transform to long format for plotting
  pivot_longer(
    cols = c(avg_sentiment, avg_bias),
    names_to = "metric",
    values_to = "score"
  ) %>%
  mutate(
    metric = case_when(
      metric == "avg_sentiment" ~ "Sentiment",
      metric == "avg_bias" ~ "Bias"
    )
  )

head(usa_combined_daily)
```

```{r}
# USA daily transformed

# Transform usa_combined_daily to wide format with date, sentiment, bias columns
usa_combined_daily_transformed <- usa_combined_daily %>%
  pivot_wider(
    names_from = metric,
    values_from = score
  ) %>%
  select(date, Sentiment, Bias) %>%
  rename(
    sentiment = Sentiment,
    bias = Bias
  )

head(usa_combined_daily_transformed)
```

```{r}
# Create monthly aggregated data from the original df_countries data
usa_combined_monthly <- df_countries %>%
  filter(location == "United States", category %in% c("health", "general", "science", "business")) %>%
  mutate(year_month = floor_date(publishedAt, "month")) %>%
  group_by(year_month) %>%
  summarise(
    sentiment = mean(compound, na.rm = TRUE),
    bias = mean(bias_score, na.rm = TRUE),
    n_articles = n(),
    .groups = 'drop'
  ) %>%
  select(year_month, sentiment, bias) %>%
  arrange(year_month)

head(usa_combined_monthly)
cat("Monthly data range:", as.character(range(usa_combined_monthly$year_month)), "\n")
cat("Number of months:", nrow(usa_combined_monthly), "\n")
```

**Fill in missing values**

```{r}
# Fill in missing data using Kalman filter

# Fill in missing data for daily df

# Create a complete date sequence to identify missing days
date_range <- seq(from = min(usa_combined_daily_transformed$date), 
                  to = max(usa_combined_daily_transformed$date), 
                  by = "day")

# Create complete daily dataframe with all dates
usa_daily_complete <- data.frame(date = date_range) %>%
  left_join(usa_combined_daily_transformed, by = "date")

# Check for missing values
cat("Missing values in daily data:\n")
cat("Sentiment:", sum(is.na(usa_daily_complete$sentiment)), "\n")
cat("Bias:", sum(is.na(usa_daily_complete$bias)), "\n")

# Apply Kalman filter to fill missing values
usa_daily_filled <- usa_daily_complete %>%
  mutate(
    sentiment_filled = na_kalman(sentiment, model = "StructTS", smooth = TRUE),
    bias_filled = na_kalman(bias, model = "StructTS", smooth = TRUE)
  ) %>%
  select(date, sentiment = sentiment_filled, bias = bias_filled)

cat("Daily data after Kalman filtering:\n")
cat("Total days:", nrow(usa_daily_filled), "\n")
cat("Missing values - Sentiment:", sum(is.na(usa_daily_filled$sentiment)), "\n")
cat("Missing values - Bias:", sum(is.na(usa_daily_filled$bias)), "\n")
head(usa_daily_filled)
```

```{r}
# Fill in missing data for monthly df

# Create a complete monthly sequence
monthly_range <- seq(from = floor_date(min(usa_combined_monthly$year_month), "month"),
                     to = floor_date(max(usa_combined_monthly$year_month), "month"),
                     by = "month")

# Create complete monthly dataframe with all months
usa_monthly_complete <- data.frame(year_month = monthly_range) %>%
  left_join(usa_combined_monthly, by = "year_month")

# Check for missing values
cat("\nMissing values in monthly data:\n")
cat("Sentiment:", sum(is.na(usa_monthly_complete$sentiment)), "\n")
cat("Bias:", sum(is.na(usa_monthly_complete$bias)), "\n")

# Apply Kalman filter to fill missing values (if any)
# Check if we actually have missing values before applying Kalman filter
if(any(is.na(usa_monthly_complete$sentiment)) || any(is.na(usa_monthly_complete$bias))) {
  cat("Applying Kalman filter to fill missing monthly values...\n")
  usa_monthly_filled <- usa_monthly_complete %>%
    mutate(
      sentiment_filled = na_kalman(sentiment, model = "StructTS", smooth = TRUE),
      bias_filled = na_kalman(bias, model = "StructTS", smooth = TRUE)
    ) %>%
    select(year_month, sentiment = sentiment_filled, bias = bias_filled)
} else {
  cat("No missing values found in monthly data - using original data...\n")
  usa_monthly_filled <- usa_monthly_complete %>%
    select(year_month, sentiment, bias)
}

cat("Monthly data after Kalman filtering:\n")
cat("Total months:", nrow(usa_monthly_filled), "\n")
cat("Missing values - Sentiment:", sum(is.na(usa_monthly_filled$sentiment)), "\n")
cat("Missing values - Bias:", sum(is.na(usa_monthly_filled$bias)), "\n")
head(usa_monthly_filled)
```

**Conversion of dataframes into time series**

```{r}
# Convert daily df into a time series

# Extract the start date for daily time series
daily_start_date <- as.Date(min(usa_daily_filled$date))
daily_start_year <- as.numeric(format(daily_start_date, "%Y"))
daily_start_yday <- as.numeric(format(daily_start_date, "%j"))

# Create daily time series objects
sentiment_daily_ts <- ts(usa_daily_filled$sentiment, 
                        start = c(daily_start_year, daily_start_yday), 
                        frequency = 365.25)

bias_daily_ts <- ts(usa_daily_filled$bias, 
                   start = c(daily_start_year, daily_start_yday), 
                   frequency = 365.25)

# Display basic info about the time series
cat("Daily Time Series Information:\n")
cat("Sentiment TS - Start:", paste(start(sentiment_daily_ts), collapse = ", "), 
    "End:", paste(end(sentiment_daily_ts), collapse = ", "), 
    "Frequency:", frequency(sentiment_daily_ts), "\n")
cat("Bias TS - Start:", paste(start(bias_daily_ts), collapse = ", "), 
    "End:", paste(end(bias_daily_ts), collapse = ", "), 
    "Frequency:", frequency(bias_daily_ts), "\n")
```

```{r}
# Convert monthly df into a time series

# Extract the start date for monthly time series
monthly_start_date <- as.Date(min(usa_monthly_filled$year_month))
monthly_start_year <- as.numeric(format(monthly_start_date, "%Y"))
monthly_start_month <- as.numeric(format(monthly_start_date, "%m"))

# Create monthly time series objects
sentiment_monthly_ts <- ts(usa_monthly_filled$sentiment, 
                          start = c(monthly_start_year, monthly_start_month), 
                          frequency = 12)

bias_monthly_ts <- ts(usa_monthly_filled$bias, 
                     start = c(monthly_start_year, monthly_start_month), 
                     frequency = 12)

# Display basic info about the time series
cat("Monthly Time Series Information:\n")
cat("Sentiment TS - Start:", paste(start(sentiment_monthly_ts), collapse = ", "), 
    "End:", paste(end(sentiment_monthly_ts), collapse = ", "), 
    "Frequency:", frequency(sentiment_monthly_ts), "\n")
cat("Bias TS - Start:", paste(start(bias_monthly_ts), collapse = ", "), 
    "End:", paste(end(bias_monthly_ts), collapse = ", "), 
    "Frequency:", frequency(bias_monthly_ts), "\n")

# Show first few values of each time series
cat("\nFirst 6 values of sentiment_monthly_ts:\n")
print(head(sentiment_monthly_ts))
cat("\nFirst 6 values of bias_monthly_ts:\n")
print(head(bias_monthly_ts))
```

**Series Visualization**

```{r}
# Plot daily sentiment

# Create date sequence for plotting
daily_dates <- usa_daily_filled$date

fig_daily_sentiment <- plot_ly(x = daily_dates, y = as.numeric(sentiment_daily_ts), 
                               type = 'scatter', mode = 'lines',
                               line = list(color = 'steelblue', width = 1.5),
                               name = 'Daily Sentiment') %>%
  layout(title = "Daily Sentiment Analysis - USA Combined Categories (Health, General, Science, Business)",
         xaxis = list(title = "Date"),
         yaxis = list(title = "Sentiment Score"),
         hovermode = 'x unified')

fig_daily_sentiment
```

```{r}
# Plot monthly sentiment

# Create date sequence for monthly plotting
monthly_dates <- usa_monthly_filled$year_month

fig_monthly_sentiment <- plot_ly(x = monthly_dates, y = as.numeric(sentiment_monthly_ts), 
                                 type = 'scatter', mode = 'lines+markers',
                                 line = list(color = 'darkblue', width = 2),
                                 marker = list(size = 6, color = 'darkblue'),
                                 name = 'Monthly Sentiment') %>%
  layout(title = "Monthly Sentiment Analysis - USA Combined Categories (Health, General, Science, Business)",
         xaxis = list(title = "Month"),
         yaxis = list(title = "Average Sentiment Score"),
         hovermode = 'x unified')

fig_monthly_sentiment
```

```{r}
# Plot daily bias

fig_daily_bias <- plot_ly(x = daily_dates, y = as.numeric(bias_daily_ts), 
                          type = 'scatter', mode = 'lines',
                          line = list(color = 'coral', width = 1.5),
                          name = 'Daily Bias') %>%
  layout(title = "Daily Bias Analysis - USA Combined Categories (Health, General, Science, Business)",
         xaxis = list(title = "Date"),
         yaxis = list(title = "Bias Score"),
         hovermode = 'x unified')

fig_daily_bias
```

```{r}
# Plot monthly bias

fig_monthly_bias <- plot_ly(x = monthly_dates, y = as.numeric(bias_monthly_ts), 
                            type = 'scatter', mode = 'lines+markers',
                            line = list(color = 'darkred', width = 2),
                            marker = list(size = 6, color = 'darkred'),
                            name = 'Monthly Bias') %>%
  layout(title = "Monthly Bias Analysis - USA Combined Categories (Health, General, Science, Business)",
         xaxis = list(title = "Month"),
         yaxis = list(title = "Average Bias Score"),
         hovermode = 'x unified')

fig_monthly_bias
```

**ADF Test for Stationarity**

```{r}
# ADF Test for stationarity of daily data

cat("=== ADF Test Results for Daily Time Series ===\n\n")

# ADF test for daily sentiment
cat("Daily Sentiment ADF Test:\n")
adf_daily_sentiment <- adf.test(sentiment_daily_ts, alternative = "stationary")
print(adf_daily_sentiment)
cat("Interpretation: ")
if(adf_daily_sentiment$p.value < 0.05) {
  cat("Series is STATIONARY (p-value < 0.05)\n")
} else {
  cat("Series is NON-STATIONARY (p-value >= 0.05)\n")
}

cat("\n" , rep("-", 50), "\n\n")

# ADF test for daily bias
cat("Daily Bias ADF Test:\n")
adf_daily_bias <- adf.test(bias_daily_ts, alternative = "stationary")
print(adf_daily_bias)
cat("Interpretation: ")
if(adf_daily_bias$p.value < 0.05) {
  cat("Series is STATIONARY (p-value < 0.05)\n")
} else {
  cat("Series is NON-STATIONARY (p-value >= 0.05)\n")
}
```

```{r}
# ADF Test for stationarity of monthly data

cat("=== ADF Test Results for Monthly Time Series ===\n\n")

# ADF test for monthly sentiment
cat("Monthly Sentiment ADF Test:\n")
adf_monthly_sentiment <- adf.test(sentiment_monthly_ts, alternative = "stationary")
print(adf_monthly_sentiment)
cat("Interpretation: ")
if(adf_monthly_sentiment$p.value < 0.05) {
  cat("Series is STATIONARY (p-value < 0.05)\n")
} else {
  cat("Series is NON-STATIONARY (p-value >= 0.05)\n")
}

cat("\n" , rep("-", 50), "\n\n")

# ADF test for monthly bias
cat("Monthly Bias ADF Test:\n")
adf_monthly_bias <- adf.test(bias_monthly_ts, alternative = "stationary")
print(adf_monthly_bias)
cat("Interpretation: ")
if(adf_monthly_bias$p.value < 0.05) {
  cat("Series is STATIONARY (p-value < 0.05)\n")
} else {
  cat("Series is NON-STATIONARY (p-value >= 0.05)\n")
}

cat("\n=== Summary of Stationarity Tests ===\n")
cat("Daily Sentiment:  ", ifelse(adf_daily_sentiment$p.value < 0.05, "STATIONARY", "NON-STATIONARY"), 
    " (p =", round(adf_daily_sentiment$p.value, 4), ")\n")
cat("Daily Bias:       ", ifelse(adf_daily_bias$p.value < 0.05, "STATIONARY", "NON-STATIONARY"), 
    " (p =", round(adf_daily_bias$p.value, 4), ")\n")
cat("Monthly Sentiment:", ifelse(adf_monthly_sentiment$p.value < 0.05, "STATIONARY", "NON-STATIONARY"), 
    " (p =", round(adf_monthly_sentiment$p.value, 4), ")\n")
cat("Monthly Bias:     ", ifelse(adf_monthly_bias$p.value < 0.05, "STATIONARY", "NON-STATIONARY"), 
    " (p =", round(adf_monthly_bias$p.value, 4), ")\n")
```

**Differencing**

```{r}
# Difference the data
diff_sentiment_daily <- diff(sentiment_daily_ts)

# Plot the differenced data
p1 <- ggplot() + 
  geom_line(aes(x = time(diff_sentiment_daily), y = as.numeric(diff_sentiment_daily)), color = "steelblue") + 
  labs(title = "Differenced Daily Sentiment - USA Combined Categories",
       x = "Time",
       y = "Differenced Sentiment Score") +
  theme_minimal()

# Display the Differenced Plot
p1

# Check stationarity of differenced series
cat("ADF Test for Differenced Daily Sentiment:\n")
adf_diff_daily_sentiment <- adf.test(diff_sentiment_daily, alternative = "stationary")
print(adf_diff_daily_sentiment)
cat("Interpretation: ")
if(adf_diff_daily_sentiment$p.value < 0.05) {
  cat("Differenced series is STATIONARY (p-value < 0.05)\n")
} else {
  cat("Differenced series is still NON-STATIONARY (p-value >= 0.05)\n")
}
```

```{r}
# No need to difference bias data as the series is stationary
```

```{r}
# Difference the data
diff_sentiment_monthly <- diff(sentiment_monthly_ts)

# Plot the differenced data
p3 <- ggplot() + 
  geom_line(aes(x = time(diff_sentiment_monthly), y = as.numeric(diff_sentiment_monthly)), color = "darkblue") + 
  labs(title = "Differenced Monthly Sentiment - USA Combined Categories",
       x = "Time",
       y = "Differenced Sentiment Score") +
  theme_minimal()

# Display the Differenced Plot
p3

# Check stationarity of differenced series
cat("ADF Test for Differenced Monthly Sentiment:\n")
adf_diff_monthly_sentiment <- adf.test(diff_sentiment_monthly, alternative = "stationary")
print(adf_diff_monthly_sentiment)
cat("Interpretation: ")
if(adf_diff_monthly_sentiment$p.value < 0.05) {
  cat("Differenced series is STATIONARY (p-value < 0.05)\n")
} else {
  cat("Differenced series is still NON-STATIONARY (p-value >= 0.05)\n")
}
```

```{r}
# Difference the data
diff_bias_monthly <- diff(bias_monthly_ts)

# Plot the differenced data
p4 <- ggplot() + 
  geom_line(aes(x = time(diff_bias_monthly), y = as.numeric(diff_bias_monthly)), color = "darkred") + 
  labs(title = "Differenced Monthly Bias - USA Combined Categories",
       x = "Time",
       y = "Differenced Bias Score") +
  theme_minimal()

# Display the Differenced Plot
p4
```

```{r}
# Check stationarity of differenced series
cat("ADF Test for Differenced Monthly Bias:\n")
adf_diff_monthly_bias <- adf.test(diff_bias_monthly, alternative = "stationary")
print(adf_diff_monthly_bias)
cat("Interpretation: ")
if(adf_diff_monthly_bias$p.value < 0.05) {
  cat("Differenced series is STATIONARY (p-value < 0.05)\n")
} else {
  cat("Differenced series is still NON-STATIONARY (p-value >= 0.05)\n")
}

# Summary of differencing results
cat("\n=== Summary of Differencing Results ===\n")
cat("Differenced Daily Sentiment:   ", ifelse(adf_diff_daily_sentiment$p.value < 0.05, "STATIONARY", "NON-STATIONARY"), 
    " (p =", round(adf_diff_daily_sentiment$p.value, 4), ")\n")
cat("Differenced Daily Bias:        ", ifelse(adf_diff_daily_bias$p.value < 0.05, "STATIONARY", "NON-STATIONARY"), 
    " (p =", round(adf_diff_daily_bias$p.value, 4), ")\n")
cat("Differenced Monthly Sentiment: ", ifelse(adf_diff_monthly_sentiment$p.value < 0.05, "STATIONARY", "NON-STATIONARY"), 
    " (p =", round(adf_diff_monthly_sentiment$p.value, 4), ")\n")
cat("Differenced Monthly Bias:      ", ifelse(adf_diff_monthly_bias$p.value < 0.05, "STATIONARY", "NON-STATIONARY"), 
    " (p =", round(adf_diff_monthly_bias$p.value, 4), ")\n")
```

**ACF Plots of differenced series**

```{r}
# Load patchwork library for combining plots
library(patchwork)

# ACF plot for differenced daily sentiment
acf_daily_sentiment <- ggAcf(diff_sentiment_daily, 50) +
  ggtitle("ACF of Differenced Daily Sentiment") +
  theme_minimal()

# ACF plot for differenced monthly sentiment
acf_monthly_sentiment <- ggAcf(diff_sentiment_monthly, 20) +
  ggtitle("ACF of Differenced Monthly Sentiment") +
  theme_minimal()

# ACF plot for differenced monthly bias
acf_monthly_bias <- ggAcf(diff_bias_monthly, 20) +
  ggtitle("ACF of Differenced Monthly Bias") +
  theme_minimal()

# PACF plot for differenced daily sentiment
pacf_daily_sentiment <- ggPacf(diff_sentiment_daily, 50) +
  ggtitle("PACF of Differenced Daily Sentiment") +
  theme_minimal()

# PACF plot for differenced monthly sentiment
pacf_monthly_sentiment <- ggPacf(diff_sentiment_monthly, 20) +
  ggtitle("PACF of Differenced Monthly Sentiment") +
  theme_minimal()

# PACF plot for differenced monthly bias
pacf_monthly_bias <- ggPacf(diff_bias_monthly, 20) +
  ggtitle("PACF of Differenced Monthly Bias") +
  theme_minimal()
```

```{r}
# Display ACF plots
cat("ACF Plots for Differenced Series:\n")
acf_daily_sentiment / acf_monthly_sentiment / acf_monthly_bias
```

```{r}
# Display PACF plots
cat("PACF Plots for Differenced Series:\n")
pacf_daily_sentiment / pacf_monthly_sentiment / pacf_monthly_bias
```

*The ACF and PACF plot of the differenced daily sentiment show remaining autocorrelation autocorrelation, hence the need for second order differencing*

*The ACF and PACF plot of the differenced monthly sentiment and bias show that most autocorrelations fall within the significance bounds after the first few lags*

**Second order differencing for daily sentiment**

```{r}
# Second-order differencing for daily sentiment
diff2_sentiment_daily <- diff(diff_sentiment_daily)

# Plot the second-order differenced data
p_diff2 <- ggplot() + 
  geom_line(aes(x = time(diff2_sentiment_daily), y = as.numeric(diff2_sentiment_daily)), color = "steelblue") + 
  labs(title = "Second-Order Differenced Daily Sentiment - USA Combined Categories",
       x = "Time",
       y = "Second-Order Differenced Sentiment Score") +
  theme_minimal()

# Display the plot
p_diff2

# Check stationarity of second-order differenced series
cat("ADF Test for Second-Order Differenced Daily Sentiment:\n")
adf_diff2_daily_sentiment <- adf.test(diff2_sentiment_daily, alternative = "stationary")
print(adf_diff2_daily_sentiment)
cat("Interpretation: ")
if(adf_diff2_daily_sentiment$p.value < 0.05) {
  cat("Second-order differenced series is STATIONARY (p-value < 0.05)\n")
} else {
  cat("Second-order differenced series is still NON-STATIONARY (p-value >= 0.05)\n")
}
```

```{r}
# ACF and PACF plots for second-order differenced daily sentiment

# ACF plot for second-order differenced daily sentiment
acf_diff2_daily_sentiment <- ggAcf(diff2_sentiment_daily, 50) +
  ggtitle("ACF of Second-Order Differenced Daily Sentiment") +
  theme_minimal()

# PACF plot for second-order differenced daily sentiment
pacf_diff2_daily_sentiment <- ggPacf(diff2_sentiment_daily, 50) +
  ggtitle("PACF of Second-Order Differenced Daily Sentiment") +
  theme_minimal()

# Display both plots side by side
cat("ACF and PACF Plots for Second-Order Differenced Daily Sentiment:\n")
acf_diff2_daily_sentiment / pacf_diff2_daily_sentiment
```

```{r}
# Comparison of first-order vs second-order differencing ACF plots
cat("Comparison: First-Order vs Second-Order Differencing ACF:\n")
acf_daily_sentiment / acf_diff2_daily_sentiment
```

*The ACF and PACF of the second order differencing show that the series may be over differenced as evidenced by potential negative correlation patterns*

**For modeling, we will be predicting monthly sentiment and bias hence the following parameter choices and modeling exercises will be focus on monthly sentiment and bias**

Parameter choices
- d = 1 since first order differencing is sufficient for stationarity
- p (AR order): From the PACF plots, we see significant spikes at lags 1 and 2 so we will test p = 0, 1, 2
- q (MA order) : From the ACF plot, we see a signficant correlation at lag 1 so we will test q = 0,1,2

**Parameter search for monthly sentiment**

```{r}
# Model 1: ARIMA (0,1,0)
model1 <- Arima(sentiment_monthly_ts, order = c(0, 1, 0), include.drift = FALSE)
summary(model1)
```

```{r}
# Model 2: ARIMA (1,1,0)
model2 <- Arima(sentiment_monthly_ts, order = c(1, 1, 0), include.drift = FALSE)
summary(model2)
```

```{r}
# Model 3: ARIMA (2,1,0)
model3 <- Arima(sentiment_monthly_ts, order = c(2, 1, 0), include.drift = FALSE)
summary(model3)
```

```{r}
# Model 4: ARIMA (0,1,1)
model4 <- Arima(sentiment_monthly_ts, order = c(0, 1, 1), include.drift = FALSE)
summary(model4)
```

```{r}
# Model 5: ARIMA (0,1,2)
model5 <- Arima(sentiment_monthly_ts, order = c(0, 1, 2), include.drift = FALSE)
summary(model5)
```

```{r}
# Model 6: ARIMA (1,1,1)
model6 <- Arima(sentiment_monthly_ts, order = c(1, 1, 1), include.drift = FALSE)
summary(model6)
```

```{r}
# Model 7: ARIMA (1,1,2)
model7 <- Arima(sentiment_monthly_ts, order = c(1, 1, 2), include.drift = FALSE)
summary(model7)
```

```{r}
# Model 8: ARIMA (2,1,2)
model8 <- Arima(sentiment_monthly_ts, order = c(2, 1, 2), include.drift = FALSE)
summary(model8)
```

*Model 1 : ARIMA (0,1,0) has the lowest AIC indicating to be the best model according to this manual selection process*

**Loop based approach: Breaking terminal in VS Code**

```{r}
# Parameter search for monthly sentiment

# Load necessary libraries
library(knitr)

# Define parameter ranges based on ACF/PACF analysis
p_range <- 0:2  # AR order: test p = 0, 1, 2 based on PACF analysis
d <- 1          # Differencing: d = 1 since first order differencing is sufficient
q_range <- 0:2  # MA order: test q = 0, 1, 2 based on ACF analysis

# Calculate total combinations
n_combinations <- length(p_range) * length(q_range)
cat("Testing", n_combinations, "ARIMA model combinations for monthly sentiment...\n\n")

# Create an empty matrix to store results
results_matrix <- matrix(NA, nrow = n_combinations, ncol = 6)

```

```{r}
# Initialize index for matrix row
i <- 1

cat("Starting ARIMA parameter search...\n")

# Loop through combinations of ARIMA model parameters
for (p in p_range) {
  for (q in q_range) {
    cat("Testing ARIMA(", p, ",", d, ",", q, ")... ")
    
    # Try fit ARIMA model with comprehensive error handling
    tryCatch({
      # Check if the data has enough observations
      if(length(sentiment_monthly_ts) < (p + q + d + 1)) {
        cat("INSUFFICIENT DATA\n")
        results_matrix[i, ] <- c(p, d, q, NA, NA, NA)
      } else {
        # Fit ARIMA model with specified (p,d,q) for monthly sentiment
        model <- Arima(sentiment_monthly_ts, order = c(p, d, q), include.drift = FALSE)
        
        # Store model parameters and AIC/BIC/AICc values in matrix
        results_matrix[i, ] <- c(p, d, q, model$aic, model$bic, model$aicc)
        
        # Print success
        cat("SUCCESS - AIC:", round(model$aic, 3), "\n")
      }
      
    }, warning = function(w) {
      cat("WARNING:", w$message, "\n")
      results_matrix[i, ] <- c(p, d, q, NA, NA, NA)
    }, error = function(e) {
      # If model fails to fit, store NA values and print error
      cat("FAILED -", e$message, "\n")
      results_matrix[i, ] <- c(p, d, q, NA, NA, NA)
    })

    # Increment row index
    i <- i + 1
    
    # Add a small delay to prevent overwhelming the system
    Sys.sleep(0.1)
  }
}

cat("\nParameter search completed.\n")

# Convert matrix to data frame
results_df <- as.data.frame(results_matrix)
colnames(results_df) <- c("p", "d", "q", "AIC", "BIC", "AICc")

# Remove rows with NA values 
results_df_clean <- results_df[complete.cases(results_df), ]

cat("Successfully fitted models:", nrow(results_df_clean), "out of", nrow(results_df), "\n")

if(nrow(results_df_clean) > 0) {
  # Find the row with the lowest AIC, BIC, and AICc
  best_aic_row <- which.min(results_df_clean$AIC)
  best_bic_row <- which.min(results_df_clean$BIC)
  best_aicc_row <- which.min(results_df_clean$AICc)
  
  # Display summary of best models
  cat("\n=== Best Model Selection Results ===\n")
  cat("Best model by AIC: ARIMA(", results_df_clean[best_aic_row, "p"], ",", 
      results_df_clean[best_aic_row, "d"], ",", results_df_clean[best_aic_row, "q"], 
      ") - AIC =", round(results_df_clean[best_aic_row, "AIC"], 3), "\n")
  cat("Best model by BIC: ARIMA(", results_df_clean[best_bic_row, "p"], ",", 
      results_df_clean[best_bic_row, "d"], ",", results_df_clean[best_bic_row, "q"], 
      ") - BIC =", round(results_df_clean[best_bic_row, "BIC"], 3), "\n")
  cat("Best model by AICc: ARIMA(", results_df_clean[best_aicc_row, "p"], ",", 
      results_df_clean[best_aicc_row, "d"], ",", results_df_clean[best_aicc_row, "q"], 
      ") - AICc =", round(results_df_clean[best_aicc_row, "AICc"], 3), "\n")
  
  # Display the results table with formatting
  cat("\n")
  knitr::kable(results_df_clean, 
               align = 'c', 
               caption = "ARIMA Model Comparison for Monthly Sentiment",
               digits = 3)
} else {
  cat("ERROR: No models were successfully fitted. Check your data.\n")
  print(summary(sentiment_monthly_ts))
}
```

**Auto selection of best model**

```{r}
auto_sentiment_monthly.model <- auto.arima(sentiment_monthly_ts, seasonal = FALSE)
cat("Auto-selected model:", auto_sentiment_monthly.model$arma[1], auto_sentiment_monthly.model$arma[6], auto_sentiment_monthly.model$arma[2])
```

```{r}
summary(auto_sentiment_monthly.model)
```

*The auto selected model is ARIMA(1,0,1)*

**Model diagnostics to get best best model**

Manual model:

```{r}
model_manual_sentiment <- capture.output(sarima(sentiment_monthly_ts, 0,1,0))
```

- The residual plot show consistent fluctuation around zero and but no constant variation which shows that the model might need improvement slightly
- The ACF plot does not reveal any significant correlations 
- The Q-Q plot indicates that the residuals follow a near-normal distribution, with minor deviations at the tails, which is typical in time series data
- The Ljung-Box Test p-values remain mostly above the 0.05 significance level, implying no significant autocorrelations left in the residuals and concluding that the model is adequate

```{r}
model_auto_sentiment <- capture.output(sarima(sentiment_monthly_ts, 1,0,1))
```

- The residual plot does not show consistent fluctuation around zero and no constant variation which shows that the model might need improvement
- The ACF plot does not reveal any significant correlations 
- The Q-Q plot indicates that the residuals follow a near-normal distribution, with major deviations at the tails, 
- The Ljung-Box Test p-values remain mostly above the 0.05 significance level, implying  no significant autocorrelations left in the residuals and concluding that the model might be adequate

**Both models perform similarly, with the residual, acf, and normal q-q plot almost identical. The only different plots are the Ljung-Box statistic plot with the automatic plot producing p-values that are significantly > 0.05**

# Benchmark methods and models


```{r}
# Define model orders for comparison
manual_order <- c(0, 1, 0)
auto_order <- c(1, 0, 1)
```

```{r}
# Split data for evaluation (use last 6 observations for testing)
n <- length(sentiment_monthly_ts)
train_data <- sentiment_monthly_ts[1:(n-6)]
test_data <- sentiment_monthly_ts[(n-5):n]

# Fit manual model on training data
manual_arima <- Arima(train_data, order = manual_order, include.drift = FALSE)
manual_forecast <- forecast(manual_arima, h = 6)

# Fit automatic model on training data
auto_arima <- Arima(train_data, order = auto_order, include.drift = FALSE)
auto_forecast <- forecast(auto_arima, h = 6)
```


```{r}
# Benchmark methods
# Naive method
naive_forecast <- naive(train_data, h = 6)

# Mean method  
mean_forecast <- meanf(train_data, h = 6)

# Random walk with drift
rwf_forecast <- rwf(train_data, h = 6, drift = TRUE)

# Calculate accuracy measures
manual_accuracy <- accuracy(manual_forecast, test_data)
auto_accuracy <- accuracy(auto_forecast, test_data)
naive_accuracy <- accuracy(naive_forecast, test_data)
mean_accuracy <- accuracy(mean_forecast, test_data)
rwf_accuracy <- accuracy(rwf_forecast, test_data)

# Create comparison table
comparison <- data.frame(
  Method = c("Manual ARIMA", "Auto ARIMA", "Naive", "Mean", "RWF with Drift"),
  MAE = c(manual_accuracy[2,3], auto_accuracy[2,3], naive_accuracy[2,3], mean_accuracy[2,3], rwf_accuracy[2,3]),
  RMSE = c(manual_accuracy[2,2], auto_accuracy[2,2], naive_accuracy[2,2], mean_accuracy[2,2], rwf_accuracy[2,2]),
  MAPE = c(manual_accuracy[2,5], auto_accuracy[2,5], naive_accuracy[2,5], mean_accuracy[2,5], rwf_accuracy[2,5])
)
```

```{r}
knitr::kable(comparison, digits = 3, caption = "Model Performance Comparison")
```

*The random walk with drift has a lower MAE, RMSE, and MAPE, outperforming both the Manual ARIMA and Auto Arima*

**Plot forecasts with manual arima as chosen model to include with baseline models as it outperforms auto arima**


```{r}
# Forecast visualization

# Fit the best manual model on full data
best_manual_fit <- Arima(sentiment_monthly_ts, order = manual_order, include.drift = FALSE)

# Create forecast plot comparing all methods
autoplot(sentiment_monthly_ts) +
  autolayer(meanf(sentiment_monthly_ts, h = 60), series = "Mean", PI = FALSE) +
  autolayer(naive(sentiment_monthly_ts, h = 60), series = "Naïve", PI = FALSE) +
  autolayer(rwf(sentiment_monthly_ts, drift = TRUE, h = 60), series = "RWF with Drift", PI = FALSE) +
  autolayer(forecast(best_manual_fit, h = 60), series = "Manual ARIMA", PI = FALSE) +
  ggtitle("Sentiment Forecast Comparison") +
  xlab("Time") + ylab("Sentiment (health, science, and general category)") +
  guides(colour = guide_legend(title = "Forecast Methods")) +
  theme_minimal()
```